---
title: "apply-samples"
author: "David Parker"
date: "Thursday, October 16, 2014"
output: html_document
---
<http://rollingyours.wordpress.com/2014/05/15/the-apply-command-101/>

The apply command 101
Posted: May 15, 2014 in Performance, R programming apply lapply tapply    

I’ll start with the apply command which expects a matrix as input. Depending
on what function you specify when using the apply command, you will get back
either a vector or a matrix.

```{r}
set.seed(1)  # Makes the call to rnorm generate the same numbers every time
( mymat <- matrix(round(rnorm(16,10),2),4,4) )

```


Let’s say that we want to take the mean of each column. Another way to say
this is that we want to apply the mean function to each column in the matrix.  
This is the way to start thinking about things like this in R. But first, how
might we do this if we don’t know anything about the apply command ?  

```{r}
mean(mymat[,1])

mean(mymat[,2])

mean(mymat[,3])

mean(mymat[,4])
```

Or we could stash the means into a vector if we need to use that information elsewhere  

```{r}
( mymatcolmean <- c(mean(mymat[,1]),mean(mymat[,2]),mean(mymat[,3]),mean(mymat[,4])) )
```

Well, if you have experience with programming languages like C, FORTRAN, Java,
or Perl then you probably know how to use a for-loop structure to solve this
problem. R supports for-loops too but it also offers some ways to avoid using
them.  

```{r}
retvec <- vector()
for (ii in 1:ncol(mymat)) {
    retvec[ii] = mean(mymat[,ii])
}
retvec
```


We could even put this into a function for later use in case we need it.  

```{r}
myloop <- function(somemat) {
    retvec <- vector()
    length(retvec) <- ncol(somemat)
    for (ii in 1:ncol(somemat)) {
        retvec[ii] <- mean(somemat[,ii])
    }
    return(retvec)
}

myloop(mymat)
```

This will now work for any matrix but of course it is specific to the columns of the matrix.  

```{r}
set.seed(1)

newmat <- matrix(rnorm(100),10,10)

myloop(newmat)
```

So this would work but what if we wanted to take the mean of the rows instead
of the columns ? We could do that but we then have to rewrite our function.
And, while we are at it, in order to make our function more general what about
providing an argument that indicates whether we want to operate on the rows or
the columns ? I’ll use “1” to indicate rows and “2” to indicate columns.  

```{r}
myloop <- function(somemat,rc=2) {
    retvec <- vector()
    length(retvec) <- ncol(somemat)
    
    # Check to see if we are taking the mean of the columns or rows
    # 1 indicates rows, 2 indicates columns
    
    if (rc == 2) {
        for (ii in 1:ncol(somemat)) {
            retvec[ii] <- mean(somemat[,ii])
        }
    } else {
        for (ii in 1:nrow(somemat)) {
            retvec[ii] <- mean(somemat[ii,])
        }
    }
    return(retvec)
}
```

Okay let's make sure it works.   

```{r}
myloop(mymat,2)   # This is correct

myloop(mymat,1)   # This is correct

mean(mymat[1,])  # A spot check
```

Doing it the easy way
Well okay that’s good but what if we then wanted our myloop function to
accommodate a function other than mean ? Like quantile, sum, fivenum, or even
a function of our own design ? We could add another argument to our function
to let the user specify the name of a function to be applied to the rows or
columns. But before we do more work please consider that R already has
something that will do the job for us – the apply command.  

```{r}
apply(mymat, 2, mean)

apply(mymat, 1, mean)
```

See how much easier that is than writing our own looping function ? It has
been my observation that those well versed in traditional programming
languages have a bigger problem getting used to the apply function than
newcomers simply because experienced programmers are more accustomed to
writing their own summary functions. They just dive in and start coding. But R
short circuits this approach by providing the apply family of commands. Note
also that we can substitute in any function we want to.  

```{r}
apply(mymat,2,class)  # What class do the columns belong to ?

apply(mymat,2,sum)    # Get the sum of all the columns

apply(mymat,1,range)  # Get the range of all rows

apply(mymat,2,fivenum) # Get the fivenum summary for each column

rowMeans(mymat)     # Equivalent to apply(mymat,1,mean)

colMeans(mymat)     # Equivalent to apply(mymat,2,mean)

rowSums(mymat)      # Equivalent to apply(mymat,1,sum)

colSums(mymat)      # Equivalent to apply(mymat,2,sum)
```

[...cont...](http://rollingyours.wordpress.com/2014/05/15/the-apply-command-101/)