---
title: "ggvis"
author: "David Parker"
date: "Tuesday, October 14, 2014"
output: html_document
---

###Installing ggvis  

```{r ggvis_install, eval=FALSE}
# Install ggvis
install.packages("ggvis")
```

##Introducing ggvis 0.4  

```{r chickwts}
library(ggvis)
chickwts %>% 
    ggvis(~feed, ~weight) %>% 
    layer_boxplots()
``` 

---  

###Scatterplots  
Examine the effect of weight on mpg using mtcars data  

```{r carsdf}
library(dplyr)
carsdf <- tbl_df(mtcars)
carsdf
carsdf %>% ggvis(~wt, ~mpg) %>% layer_points()
```

Some tweaks  

```{r carsdftweak}
carsdf %>% ggvis(~wt, ~mpg) %>% 
    layer_points(size := 25, shape := "diamond", stroke := "red", fill := NA)

```

---  

###Regression lines  

```{r mtcars_regress}
mtcars %>%
    ggvis(~wt, ~mpg) %>%
    layer_smooths() %>%
    layer_points()
```  


With a linear model, and 95% confidence interval for the model:  

```{r mtcars_lm}
mtcars %>% 
  ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  layer_model_predictions(model = "lm", se = TRUE)

```

---  

###Scatterplots with grouping  
Coloring points by variables   

```{r mtcars_scatter}
mtcars %>%
    ggvis(~wt, ~mpg) %>%
    layer_points(fill = ~factor(cyl))
```  

```
Coloring points, and adding a smoother for each group. The grouping variable (which is applied before the transform_smooth is calculated) must be specified with group_by():  
```

```{r mtcars_groupby}
mtcars %>% 
  ggvis(~wt, ~mpg, fill = ~factor(cyl)) %>% 
  layer_points() %>% 
  group_by(cyl) %>% 
  layer_model_predictions(model = "lm")
``` 

---  

###Bar graphs  
Using the `pressure` dataset  

```{r pressure}
head(pressure)
```  

When the variable onn the X axis is continuous  

```{r pressure_bar}
pressure %>%
    ggvis(~temperature, ~pressure) %>%
    layer_bars()
```

Specifying the width of the bars:  

```{r pressure_barwidth}
pressure %>%
    ggvis(~temperature, ~pressure) %>%
    layer_bars(width = 15)
```


When the X axis variable is categorical:  

```{r pressure_barfactor}
pressure2 <- pressure %>% mutate(temperature = factor(temperature))
head(pressure2)    
pressure2 %>%
    ggvis(~temperature, ~pressure) %>%
    layer_bars(width = 0.5)
```

