---
title: "dplyr-intro"
author: "David Parker"
date: "Wednesday, September 24, 2014"
output:
  html_document:
    keep_md: yes
---

tutorial from [R Bloggers](http://rpubs.com/justmarkham/dplyr-tutorial)


```{r}
# library(plyr)
# detach(package:plyr)
suppressMessages( library(dplyr) )
library(hflights)
data(hflights)
head(hflights)
```

* `tbl_df` creates a local data frame
* This is a wrapper for data frame, prevents accidental printing

```{r}
flights <- tbl_df(hflights)

# samply table data frame prnting
flights

print(flights, n=20)
data.frame(head(flights))
# or
head(hflights)
```

## verb 1  
## filter: Keep matching critera

```{r}
# First, base R approach
flights[flights$Month == 1 & flights$DayofMonth == 1, ]
```

Dplyr filter

```{r}
# dplyr approach
# comma defaults to &
filter(flights, Month == 1 & DayofMonth == 1)

# use | for or
filter(flights, UniqueCarrier == "AA" | UniqueCarrier == "UA")

# %in% works as well
filter(flights, UniqueCarrier %in% c("AA", "UA") )

```

## Verb 2  
## select: Pck columns by name  

```{r}
# base R
flights[, c("DepTime", "ArrTime", "FlightNum")]
```  

```{r}
# dplyr
select(flights, DepTime, ArrTime, FlightNum)
```  

```{r}
# use colon ro select range of cols, and use `contains` to match by partial name
# can use `starts_with`, `ends_with`, and `matches` (regex)
select(flights, Year:DayofMonth, contains("Taxi"), contains("Delay"))
```

## "Chaining" or "Pipelining"  
* Normal method is nesting operations
* With %>% can write commands in natural order

```{r}
# nesting method
# select UniqueCarrier and DeDelay cols and filter for delays > 60 mins
filter(select(flights, UniqueCarrier, DepDelay), DepDelay > 60)
```

```{r}
# chaining method 
flights %>% 
    select(UniqueCarrier, DepDelay) %>% 
    filter(DepDelay > 60)
```  

```{r}
# %>% comes from magrittr, used with other commands
# create 2 vectors and calc Euclidean dist between them
(x1 <- 1:5); (x2 <- 2:6)
sqrt(sum( (x1 - x2)^2 ) )
```

```{r}
# chaining method
(x1 - x2)^2 %>% sum() %>% sqrt()
```

## verb 3
## arrange: Reorder rows (sort)

```{r}
# base R: select UniqueCarrier and DepDelay, sort by DepDelay
flights[order(flights$DepDelay), c("UniqueCarrier", "DepDelay")]
```

```{r}
# dplyr approach
flights %>% 
    select(UniqueCarrier, DepDelay) %>%
    arrange(DepDelay)
```

```{r}
# use `desc` for descending
flights %>% 
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay) )
```  

## verb 4
## mutate: Add new variables (transform)
* Create new variables from functions of existing
  
```{r}
# base R approach to create  a new variable, Speed
flights$Speed <- flights$Distance / flights$AirTime*60
flights[, c("Distance", "AirTime", "Speed")]
```

```{r}
# dplyr approach (prints but does not modify flights ldf)
flights %>%
    select(Distance, AirTime) %>%
    mutate(Speed = Distance / AirTime*60)
```

```{r}
# to store new variable
flights2 <- flights %>%
    select(Distance, AirTime) %>%
    mutate(Speed = Distance / AirTime*60)

flights <- tbl_df(hflights)
```  

## verb 5
## summarise: Reduce variables to values  
* Primarily useful with data that has been grouped by
* `group_by` creates the groups to be operated on
* `summarise` uses the provided aggregation function to summarise each group

```{r}
# base R approach to calc avg arrival delay to each dest
head(with(flights, tapply(ArrDelay, Dest, mean, na.rm = TRUE)))
head(aggregate(ArrDelay ~ Dest, flights, mean))
```  

```{r}
# dplyr approach
flights %>%
    group_by(Dest) %>%
    summarise(avg_delay = mean(ArrDelay, na.rm = TRUE))
# same as
flights %>%
    select(ArrDelay, Dest) %>%
    group_by(Dest) %>%
    summarise(avg_delay = mean(ArrDelay, na.rm = TRUE))
```  
* `summarise_each` allows for performing same summary function to mult cols
* Note: `mutate_each` is also available
```{r}
# for each carrier, cal pct of flights cancelled or diverted
# Cancelled & Diverted are binary, 0 or 1
flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs(mean), Cancelled, Diverted)
```

```{r}
# for each carrier, calc min & max arrival & departure delays
flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs (min(., na.rm=T), max(., na.rm=T) ), matches("Delay"))
```  
* Helper function `n()` counts number of rows in a  group
* Helper function `n_distinct()` counts unique items in vector
```{r}
# for each day in year, count tot number of flights, sort in desc order
flights %>%
    group_by(Month, DayofMonth) %>%
    summarise(flight_count = n()) %>%
    arrange(desc(flight_count))
```

```{r}
# rewrite above more simply with `tally()`
flights %>%
    group_by(Month, DayofMonth) %>%
    tally(sort = TRUE)
```
* Grouping can be useful without summarising
```{r}
# for each dest, show the number of cancelled and notcancelled flights
flights %>%
    group_by(Dest) %>%
    select(Cancelled) %>%
    table() %>%
    head()
```

## Window Functions
* Aggregation function, like `mean`, takes n inputs and returns 1 value
* [Window function](http://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html) returns n values
```{r}
# for each carrier, alculate which two days of the year they had their longest departure delays
#  note: smallest (not largest) value is ranked as 1, so you have to use `desc` to rank by largest value
flights %>%
    group_by(UniqueCarrier) %>%
    select(Month, DayofMonth, DepDelay) %>%
    filter(min_rank(desc(DepDelay) ) <= 2 ) %>%
    arrange(UniqueCarrier, desc(DepDelay))
```   

```{r}
# rewrite more simply with the `top_n` function
flights %>%
    group_by(UniqueCarrier) %>%
    select(Month, DayofMonth, DepDelay) %>%
    top_n(2) %>%
    arrange(UniqueCarrier, desc(DepDelay))
```  

```{r}
# for each month, calculate the number of flights and the change from the previous month
flights %>%
    group_by(Month) %>%
    summarise(flight_count = n()) %>%
    mutate(change = flight_count - lag(flight_count))
```

```{r}
# rewrite more simply with the `tally` function
flights %>%
    group_by(Month) %>%
    tally() %>%
    mutate(change = n - lag(n))
```

## Other Useful Convenience Function

```{r}
# randomly sample a fixed number of rows, without replacement
flights %>% sample_n(5)
```

```{r}
# randomly sample a fraction of rows, with replacement
flights %>% sample_frac(0.25, replace = TRUE)
```

```{r}
# base R approach to view the structure of an object
str(flights)
```

```{r}
# dplyr approach, better formatting and adapts for screen width
glimpse(flights)
```

## Connecting to Databases
* dply can connect to a db as if it were loaded into a df
* Uses the same syntax for db & df
* Only generates Select stmts (cannot modify data in db)
* Currently supports SQLite, PostgreSQL/Redshift, MySQL/MariaDB, BigQuery, MonetDB
* Examples below based upon SQLite db containing hflights data
    * Instructions are in [database vignette](http://cran.r-project.org/web/packages/dplyr/vignettes/databases.html)

```{r}
my_db <- src_sqlite("my_db.sqlite3", create = T)
```

```{r}
# connect to an SQLite database containing the hflights data
flights_sqlite <- copy_to(my_db, hflights, temporary = FALSE)
```


```{r}
# connect to the "hflights" table in that database
flights_tbl <- tbl(my_db, "hflights")

# example query with our data frame
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay))
```

```{r}
# identical query using the database
flights_tbl %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay))
```

```{r}
# ask dplyr for the SQL commands
flights_tbl %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay)) %>%
    explain()
```

