---
title: "dplyr-03"
author: "David Parker"
date: "Monday, October 13, 2014"
output: html_document
---


##[What's new](http://blog.rstudio.org/2014/10/13/dplyr-0-3-2/)

Four new features:  
* Four new verbs: distinct(), slice(), rename(), and transmute()  
* Three new helper functions: between, count(), and data_frame()  
* More flexible join specifications  
* Support for row based operations  

###New verbs  
distinct() returns distinct(unique) rows of a table:  
```{r}
library("dplyr")
library("nycflights13")
data(flights)
# Find all origin-destination flight pairs
flights %>%
    select(origin, dest) %>%
    distinct()
```

slice() allows you to select rows by position. It includes positive integers and drops negative integers:  

```{r}
# Get the first flight to each destination
flights %>%
    group_by(dest) %>%
    slice(1)
```  

transmute() and rename() are variants of mutate() and select(). Transmute drops all columns  that you don't specifically mention, rename keeps all columns that you don't specifically mention. They complete this table:  

```
                           | Drop others     | Keep others     |  
---------------------------|-----------------|-----------------|
Rename & reorder variables | select()        | rename()        |  
Compute new variables      | transmute()     | mutate()        |  
```  

###New helpers  
data_frame(),  contributed by [Kevin Ushey](https://github.com/kevinushey), is a nice way to create data frames:  
* It never changes the type of inputs (i.e. no more stringAsFactors = FALSE!)  

```{r}
data.frame(x = letters) %>% sapply(class)
    
data_frame(x = letters) %>% sapply(class)
```  

* Or the names of variables:  

```{r}
data.frame(`crazy name` = 1) %>% names()

data_frame(`crazy name` = 1) %>% names()
```  

* It evaluates arguments lazily and in order:  

```{r}
data_frame(x = 1:5, y = x ^ 2)
```  

* It adds tbl_df() class to output, never adds row.names(), and only recycles vectors of length 1.

The count() function wraps up the common combination of group_by() and summarise():  

```{r}
# How many flights to each destination?
flights %>% count(dest)

# Which planes flew the most?
flights %>% count(tailnum, sort = TRUE)

# What's the total carrying capacity of planes by year of purchas
planes %>% count(year, wt = seats)
```

###Better joins  
You can now join by different variable in each table:

```{r}
narrow <- flights %>% select(origin, dest, year:day)

# Add destination airport metadata 
narrow %>%  left_join(airports, c("dest" = "faa"))

# Add origin airport metadata
narrow %>% left_join(airports, c("origin" = "faa"))
```

(right_join() and outer_join() implementations are planned for dplyr 0.4.)  
###Set operations  
You can use intersect(), union() and setdiff() with data frames, data tables, and databases:  

```{r}
jfk_planes <- flights %>%
    filter(origin == "JFK") %>%
    select(tailnum) %>%
    distinct()  
lga_planes <- flights %>%
    filter(origin == "LGA") %>%
    select(tailnum) %>%
    distinct()

# Planes that fly out of either JFK or LGA
nrow(union(jfk_planes, lga_planes))

# Planes that fly out of both JFK or LGA
nrow(intersect(jfk_planes, lga_planes))

# Planes that fly out of JFK but not LGA
nrow(setdiff(jfk_planes, lga_planes))

```  

###Programming with dplyr  
You can now program with dplyr – every function that uses non-standard evaluation (NSE) also has a standard evaluation (SE) twin that ends in _. For example, the SE version of filter() is called filter_(). The SE version of each function has similar arguments, but they must be explicitly “quoted”. Usually the best way to do this is to use ~:  

```{r}
airport <- "ANC"
# NSE version
filter(flights, dest == airport)

# Equivalent SE version
criteria <- ~dest == airport
filter_(flights, criteria)
```  

To learn more, read the [Non-standard evaluation](http://cran.r-project.org/web/packages/dplyr/vignettes/nse.html) vignette. This new approach is powered by the [lazyeval package](https://github.com/hadley/lazyeval) which provides all the tools needed to implement NSE consistently and correctly. I now understand how to implement NSE consistently and correctly, and I’ll be using the same approach everywhere.  



